### 2024-04-20 MSRI-16-1989-79.Rmd
more of 1

::: {.rmenv #chunk-1.15 title="1.15 Variant" latex="{1.15 Variant}"}
Let $P$ be a finite set of prime numbers.
The category of *smooth realisation systems on $\Spec(\ZZ)\setminus P$* is defined as in \ref{minus-three-points-1.9}, taking $P$ to be the exceptional set $S$ in [(AM3)](am3), and replacing "almost all $p$" in [(M4)](m4) and [(AM4)](am4) by "all $p\not\in P$".
This treatment of crystalline structures is not satisfying, cf. [(1.7)](#chunk-1.7).
The category \ref{minus-three-points-1.9} of realisation systems on $\Spec(\QQ)$ is the inductive limit of these categories for $P$ growing larger and larger.

Instead of saying "smooth over $\Spec(\ZZ)\setminus P$", we also say "of good reduction outside of $P$".
This terminology is erroneous in that the categories in question are not subcategories of the category of realisation systems on $\Spec(\QQ)$ (cf. [(1.7)](#chunk-1.7) again).

\oldpage{100}
In the language of [(1.14)](#chunk-1.14), in [(M'3)](mp3) we need to give $M_\ell$ as a smooth $\QQ_\ell$-sheaf on $\Spec(\ZZ[1/\ell])\setminus P$ instead of on $\Spec(\QQ)$, and we need to modify [(M'4)](mp4) like [(M4)](m4).

The objects [(1.15)](#chunk-1.15) belonging to the subcategory generated by the objects of geometric origin (cf. [(1.11)](#chunk-1.11)) will be called *smooth (mixed) motives on $\Spec(\ZZ)\setminus P$*.
We hope that this gives a full subcategory of the category of motives on $\Spec(\QQ)$.

We would like to have a notion of smooth motive on $S$ for more general base spaces than $\Spec(\ZZ)\setminus P$.
Our methods, where the Betti realisation plays a central role, require that $S_\QQ$ be dense in $S$.
We will outline a provisional definition of smooth realisation systems on $S$, for $S$ smooth over $\Spec(\ZZ)$.
The case where $S$ is open in the spectrum of the ring of integers of a number field can be dealt with using natural modifications of \ref{minus-three-points-1.4} and [(1.15)](#chunk-1.15).
For $S$ finite and étale over an open of $\Spec(\ZZ)$, we can also reduce to [(1.15)](#chunk-1.15): see [(1.17)](#chunk-1.17).
:::

:::: {.rmenv #chunk-1.16 title="1.16 Example" latex="{1.16 Example}"}
(The motivic $\HH^0$ of the spectrum of a number field).

Let $E$ be a finite extension of $\QQ$, unramified outside of a finite set $P$ of prime numbers.
We are going to expand on [(1.6)](#chunk-1.6) for $\Spec(E)$ and, more precisely, define a realisation system $A\coloneqq\HH_\mot^0(\Spec(E))$ that is smooth over $\Spec(\ZZ)\setminus P$.
The motive $A$ is of Hodge type $(0,0)$.
This determines $W$ and $F$.
We have that $A_\DR=E$, viewed as a vector space over $\QQ$.

Let $\Hom(E,\CC)=\Spec(E)(\CC)$ be the set of homomorphisms (which are automatically embeddings) from $E$ to $\CC$.
We have that $A_\BB=\QQ^{\Hom(E,\CC)}$, with $F_\infty$ induced by the complex conjugation of $\CC$.
The comparison isomorphism $\DR/\BB$ from $A_\BB\otimes\CC=\CC^{\Hom(E,\CC)}$ to $A_\DR\otimes\CC=E\otimes\CC$ is the $\CC$-linear extension of the map
$$
  \begin{aligned}
    E &\longrightarrow \CC^{\Hom(E,\CC)}
  \\e &\longmapsto (\sigma\mapsto\sigma(e)).
  \end{aligned}
$$

Since every embedding of $E$ into $\CC$ factors through $\bar{\QQ}$, $\Gal(\bar{\QQ}/\QQ)$ acts on $\Hom(E,\CC)$,
This action induces the Galois action on $A_\AAf\coloneqq A_\BB\otimes\AAf$ (cf. [(1.5.ii)](#chunk-1.5)).

For $p\not\in P$, and $v$ running over the places of $E$ over $p$, we have that $E\otimes\QQ_p=\prod E_v$, and $\phi_p$ is the automorphism of $E\otimes\QQ_p$ that induces on each $E_v$ the unique lift of the Frobenius $x\mapsto x^p$ of the residue field (cf. [(1.5.iii)](#chunk-1.5)).

\oldpage{101}The motive $A$ is endowed with a product
$$
  A\otimes A\to A
\startverb\tag{1.16.1}\stopverb
$$
namely the cup product, which makes $A$ a commutative ring with unit in the Tannakian category of motives (cf. [(5.3)](#chunk-5.3)).
On $A_\DR=E$, it is the product.
On $A_\BB$, it is given by $(q_1(\sigma))(q_2(\sigma))=(q_1(\sigma)q_2(\sigma))$.
::::

::: {.rmenv #chunk-1.17 title="1.17 Definition" latex="{1.17 Definition}"}
Let $\OO$ be the ring of $P$-integers of $E$.
With the above notation, a _smooth realisation system on $\Spec(\OO)$_ is a realisation system $N$ on $\Spec(\ZZ)\setminus P$ (cf. [(1.15)](#chunk-1.15)) endowed with the structure of an $A$-module $A\otimes N\to N$ (cf. [(5.3)](#chunk-5.3)) over $A\coloneqq\HH_\mot^0(\Spec(E))$.
:::

Similarly for "motive" and "over $\Spec(E)$" (taking the limit over $P$).

:::: {.rmenv #chunk-1.18 title="1.18" latex="{1.18}"}
We will now show how a realisation system $N$, smooth over $S=\Spec(\OO)$ (as in [(1.17)](#chunk-1.17)), can be described in terms of a *realisation system over $E$*, of the following type.

@@

We have $N_\BB=\bigoplus M_\sigma$, with the evident structure of a module over $A_\BB=\QQ^{\Hom(E,\CC)}$, and $F_\infty$ is the sum of the $F_\infty$.
As in [(1.14)](#chunk-1.14), we have a variant:

@@

::: {.trmenv #Mp3E title="(M'3)E" latex="{(M'3)\textsubscript{E}}"}
(cf. [(M3)](m3), [(M9)](m9)) An $\AAf$-sheaf $M_\AAf$ on $\Spec(E)$ (cf. [(1.14)](#chunk-1.14)).
:::

@@

By summing over $\sigma$ (resp. $v$), these give [(M'5)](mp5).

@@

We leave to the reader the task of translating axioms [(AM1)](am1), [(AM3)](am3), and [(AM4)](am4), which remain to be imposed, into this language (cf. [(1.14)](#chunk-1.14)).
::::

::: {.rmenv #chunk-1.19 title="1.19" latex="{1.19}"}
With the definition [(1.17)](#chunk-1.17) of smooth realisation systems over $S=\Spec(\OO)$, the functor given by forgetting the $A$-module structure, which takes values in smooth realisation systems over $\Spec(\ZZ)\setminus P$, is called the "*direct image of $S$ in $\Spec(\ZZ)\setminus P$*".
In the various realisations of [(1.18)](#chunk-1.18), it corresponds to the direct image.
Its left adjoint, $M\mapsto A\otimes M$ (cf. [(5.3)](#chunk-5.3)), is the *inverse image*.
:::

@@

- In [(AM3)](am3), we ask for each $\QQ_\ell$-sheaf $M_\ell$ induced by $M_\AAf$ to come from a smooth $\QQ_\ell$-sheaf on $S[1/\ell]$.
- In [(M'4)E](#Mp4E), for each residue field $k(v)$ of $S$, we ask for $M_\crisv$ over the field of fractions $K_v$ of Witt vectors over $k(v)$, endowed with a semi-linear $\phi_v$.
  The crystalline comparison isomorphism of [(M'5)E](#Mp5E) then becomes
  $$
    M_\DR\otimes E_v \simto M_\crisv\otimes_{K_v}E_v.
  $$
:::

@@

::: {.trmenv #Mp5S title="(M'5)S" latex="{(M'5)\textsubscript{S}}"}
Comparison isomorphisms
$$
  \begin{aligned}
    \comp_{\DR,\BB}\colon
    &M_\BB(C)\otimes C \simto (M_\DR\otimes C)^\nabla
  \\\comp_{\AAf,\BB}\colon
    &M_\BB^{(C)}\otimes\AAf \to \text{inverse image of }M_\AAf\text{ on }S(C)
  \end{aligned}
$$
that are functorial in $C$, where we denote by $M_\DR\otimes C$ the inverse image of $M_\DR$ on $S_C$, and by $(-)\nabla$ the sheaf of its horizontal sections on $S(C)$.
Instead of giving $M_\AAf$ in [(M'3)S](#Mp3S), we can only give the $\QQ_\ell$ sheaves $M_\ell$ that are deduced from it, along with the comparison isomorphisms $\comp_{\ell,\BB}$ from $M_\BB\otimes\QQ_\ell$ to the inverse image of $M_\ell$, and impose the existence of a lattice $L$ in $M_\BB(C)$ such that $L\otimes\ZZ_\ell$ corresponds to a $\ZZ_\ell$-sheaf $L_\ell$ with $L_\ell\otimes\QQ_\ell\simto M_\ell$, cf. [(1.14)](#chunk-1.14).

@@

The axioms are modified as follows.
In [(AM1)](am1), we want a variation of mixed Hodge structures;
[(AM3)](am3) becomes: $M_\ell$ comes from a smooth $\QQ_\ell$-sheaf on $S[1/\ell]$;
for [(AM4)](am4), a condition is imposed for every closed point of $S$.

@@

::: {.rmenv #chunk-1.22 title="1.22 Definition" latex="{1.22 Definition}"}
\oldpage{104}Let $E$ be a finite extension of $\QQ$.
Then a realisation system *with coefficients in $E$* is a realisation system $M$ endowed with the structure of an $E$-modules, $E\to\End(M)$.
:::

@@

::: {.rmenv #chunk-2.1 title="2.1" latex="{2.1}"}
\oldpage{105}The *Tate motive* $\ZZ(1)$ is a motive over $\Spec(\ZZ)$ [(1.15)](#chunk-1.15) with integer coefficients [(1.23)](#chunk-1.23).
Here is its description as a realisation system, in the language of \ref{minus-three-points-1.9} and [(1.14)](#chunk-1.14) (along with [(1.25)](#chunk-1.25)):

- [(M1)](m1) and [(M7)](m7): $\ZZ(1)_\BB=2\pi i\ZZ\subset\CC$ and $F_\infty=-1$.
- [(M'1)](mp1): $\ZZ(1)_\BB(C)=2\pi i\ZZ\subset C$.
- [(M2)](m2) and [(M'2)](mp2): $\ZZ(1)_\DR=\QQ(1)_\DR=\QQ$.
- [(M3)](m3) and [(M9)](m9): $\ZZ(1)_{\ZZ/n}$ is the group $\mu_n\subset\CC^\times$ of $n$-th roots of unity.
  The transition morphisms are the $\mu_n\to\mu_m\colon x\mapsto x^{n/m}$ for $m\mid n$.
  The action of $\Gal(\bar{\QQ}/\QQ)$ is its action on the $\mu_n$.
  In the variant [(M'3)](mp3), we have $\ZZ(1)_{\ZZ/n}(\bar{k})=\mu_n(\bar{k})$.

  We thus have that $\ZZ(1)_\ell=\limproj\mu_{\ell^n}(\CC)$.
- [(M4)](m4), [(M10)](m10), and [(M'4)](mp4): $\ZZ(1)_\crisp=\QQ_p$ and $\phi_p=1/p$.
- [(M5)](m5): The canonical comparison isomorphism $\comp_{\DR,\BB}$ is induced by the inclusions $\ZZ(1)_\BB=2\pi i\ZZ\subset\CC$ and $\ZZ(1)_\DR=\QQ\subset\CC$, i.e.
  $$
    \ZZ(1)_\BB\otimes\CC \simto \CC \simfrom \ZZ(1)_\DR\otimes\CC.
  $$
  The isomorphism $\comp_{\ZZ/n,\BB}$ [(1.25)](#chunk-1.25) is induced by
  $$
    \exp(x/n)\colon \ZZ(1)_\BB \to \ZZ/n(1)\subset\CC^\times.
  $$
  The isomorphism $\comp_{\crisp,\DR}$ is induced by the inclusion $\QQ\subset\QQ_p$.
- [(M'5)](mp5): Replace $\CC$ by $C$ in [(M5)](m5).
- [(M6)](m6) and [(M8)](m8): $\ZZ(1)$ is of pure weight $-2$, and $\ZZ(1)_\DR$ is of pure Hodge filtration $-1$: the Hodge type is $(-1,-1)$.
:::

@@

::: {.trmenv #M7P title="(M7)P" latex="{(M7)P}"}
\oldpage{108}The data of $F_\infty$ for $E$ is equivalent to that of an involution of $P_\BB$ that is compatible with that of $M_\BB$.
Axiom [(AM2)](am2) for $E$ is that this involution is induced by the $\CC$-antilinear involution of $P_{\BB,\CC}=\PP_{\DR,\CC}$ with fixed points $P_{\DR,\RR}$.
:::

::: {.trmenv #M8P title="(M8)P" latex="{(M8)P}"}
The data of the Hodge filtration of $E_\DR$ is equivalent to that of the intersection $F^0P_\DR$ of $F^0E_\DR$ over $P_\DR$, which is a torsor for $F^0M_\DR$: for $p>0$, $F^pE_\DR=F^pM_\DR$; for $p=0$, $F^0E_\DR$ is generated by $F^0P_\DR$; and for $p<0$, $F^pE_\DR$ is the sum of $F^0E_\DR$ and $F^pM_\DR$.
In other words, the data [(M8)](m8) for $E_\DR$ is that of a reduction of the structure group of $P_\DR$, from $M_\DR$ to $F^0M_\DR$.

The data of [(M1)](m1), [(M2)](m2), $\comp_{\DR/\BB}$, and [(M8)](m8) are equivalent to the data of a torsor $F^0P_\DR$ for $F^0M_\DR$ and, in the $M_\DR\otimes\CC=M_\BB\otimes\CC$ torsor $P_{\DR,\CC}$ which is induced by $F^0M_\DR\to M_\DR\otimes\CC$, a $\QQ$-structure $P_\BB\subset P_{\DR,\CC}$ that is an $M_\BB$ torsor.
:::

@@

We leave to the reader the task of translating axioms [(AM1)](am1) to [(AM5)](am5) for $P$.
We note only that [(AM1)](am1) and [(AM4)](am4) follow from the same axioms for $M$, and that [(AM2)](am2) and [(AM5)](am5) can be expressed at best as functoriality in $\CC$, cf. [(1.14)](#chunk-1.14).
::::

>>>>

@@

    We have $N_\BB=\bigoplus M_\sigma$, with the evident structure of a module over $A_\BB=\QQ^{\Hom(E,\CC)}$, and $F_\infty$ is the sum of the $F_\infty$.
    As in \ref{minus-three-points-chunk-1.14}, we have a variant:

@@

    ::: {.trmenv #Mp3E title="(M'3)E" latex="{(M'3)\textsubscript{E}}"}
    (cf. [(M3)](minus-three-points-m3), [(M9)](minus-three-points-m9)) An $\AAf$-sheaf $M_\AAf$ on $\Spec(E)$ (cf. \ref{minus-three-points-chunk-1.14}).
    :::

@@

    By summing over $\sigma$ (resp. $v$), these give [(M'5)](minus-three-points-mp5).

@@

    We leave to the reader the task of translating axioms [(AM1)](minus-three-points-am1), [(AM3)](minus-three-points-am3), and [(AM4)](minus-three-points-am4), which remain to be imposed, into this language (cf. \ref{minus-three-points-chunk-1.14}).
::::

::: {.rmenv #chunk-1.19 title="1.19" latex="{1.19}"}
    With the definition \ref{minus-three-points-1.17} of smooth realisation systems over $S=\Spec(\OO)$, the functor given by forgetting the $A$-module structure, which takes values in smooth realisation systems over $\Spec(\ZZ)\setminus P$, is called the "*direct image of $S$ in $\Spec(\ZZ)\setminus P$*".
    In the various realisations of [(1.18)](#chunk-1.18), it corresponds to the direct image.
    Its left adjoint, $M\mapsto A\otimes M$ (cf. \fref{minus-three-points-5.3}), is the *inverse image*.
:::

@@

    - In [(AM3)](minus-three-points-am3), we ask for each $\QQ_\ell$-sheaf $M_\ell$ induced by $M_\AAf$ to come from a smooth $\QQ_\ell$-sheaf on $S[1/\ell]$.
    - In [(M'4)E](#Mp4E), for each residue field $k(v)$ of $S$, we ask for $M_\crisv$ over the field of fractions $K_v$ of Witt vectors over $k(v)$, endowed with a semi-linear $\phi_v$.
      The crystalline comparison isomorphism of [(M'5)E](#Mp5E) then becomes
      $$
        M_\DR\otimes E_v \simto M_\crisv\otimes_{K_v}E_v.
      $$
:::

@@

    ::: {.trmenv #Mp5S title="(M'5)S" latex="{(M'5)\textsubscript{S}}"}
    Comparison isomorphisms
    $$
      \begin{aligned}
        \comp_{\DR,\BB}\colon
        &M_\BB(C)\otimes C \simto (M_\DR\otimes C)^\nabla
      \\\comp_{\AAf,\BB}\colon
        &M_\BB^{(C)}\otimes\AAf \to \text{inverse image of }M_\AAf\text{ on }S(C)
      \end{aligned}
    $$
    that are functorial in $C$, where we denote by $M_\DR\otimes C$ the inverse image of $M_\DR$ on $S_C$, and by $(-)\nabla$ the sheaf of its horizontal sections on $S(C)$.
    Instead of giving $M_\AAf$ in [(M'3)S](#Mp3S), we can only give the $\QQ_\ell$ sheaves $M_\ell$ that are deduced from it, along with the comparison isomorphisms $\comp_{\ell,\BB}$ from $M_\BB\otimes\QQ_\ell$ to the inverse image of $M_\ell$, and impose the existence of a lattice $L$ in $M_\BB(C)$ such that $L\otimes\ZZ_\ell$ corresponds to a $\ZZ_\ell$-sheaf $L_\ell$ with $L_\ell\otimes\QQ_\ell\simto M_\ell$, cf. \ref{minus-three-points-chunk-1.14}.

@@

    The axioms are modified as follows.
    In [(AM1)](minus-three-points-am1), we want a variation of mixed Hodge structures;
    [(AM3)](minus-three-points-am3) becomes: $M_\ell$ comes from a smooth $\QQ_\ell$-sheaf on $S[1/\ell]$;
    for [(AM4)](minus-three-points-am4), a condition is imposed for every closed point of $S$.

@@

::: {.rmenv #chunk-1.22 title="1.22 Definition" latex="{1.22 Definition}"}
    \oldpage{104}
    Let $E$ be a finite extension of $\QQ$.
    Then a realisation system *with coefficients in $E$* is a realisation system $M$ endowed with the structure of an $E$-modules, $E\to\End(M)$.
:::

@@

::: {.rmenv #chunk-2.1 title="2.1" latex="{2.1}"}
\oldpage{105}The *Tate motive* $\ZZ(1)$ is a motive over $\Spec(\ZZ)$ \ref{minus-three-points-1.15} with integer coefficients [(1.23)](#chunk-1.23).
Here is its description as a realisation system, in the language of \ref{minus-three-points-1.9} and \ref{minus-three-points-chunk-1.14} (along with [(1.25)](#chunk-1.25)):

- [(M1)](minus-three-points-m1) and [(M7)](minus-three-points-m7): $\ZZ(1)_\BB=2\pi i\ZZ\subset\CC$ and $F_\infty=-1$.
- [(M'1)](minus-three-points-mp1): $\ZZ(1)_\BB(C)=2\pi i\ZZ\subset C$.
- [(M2)](minus-three-points-m2) and [(M'2)](minus-three-points-mp2): $\ZZ(1)_\DR=\QQ(1)_\DR=\QQ$.
- [(M3)](minus-three-points-m3) and [(M9)](minus-three-points-m9): $\ZZ(1)_{\ZZ/n}$ is the group $\mu_n\subset\CC^\times$ of $n$-th roots of unity.
  The transition morphisms are the $\mu_n\to\mu_m\colon x\mapsto x^{n/m}$ for $m\mid n$.
  The action of $\Gal(\bar{\QQ}/\QQ)$ is its action on the $\mu_n$.
  In the variant [(M'3)](minus-three-points-mp3), we have $\ZZ(1)_{\ZZ/n}(\bar{k})=\mu_n(\bar{k})$.

  We thus have that $\ZZ(1)_\ell=\limproj\mu_{\ell^n}(\CC)$.
- [(M4)](minus-three-points-m4), [(M10)](minus-three-points-m10), and [(M'4)](minus-three-points-mp4): $\ZZ(1)_\crisp=\QQ_p$ and $\phi_p=1/p$.
- [(M5)](minus-three-points-m5): The canonical comparison isomorphism $\comp_{\DR,\BB}$ is induced by the inclusions $\ZZ(1)_\BB=2\pi i\ZZ\subset\CC$ and $\ZZ(1)_\DR=\QQ\subset\CC$, i.e.
  $$
    \ZZ(1)_\BB\otimes\CC \simto \CC \simfrom \ZZ(1)_\DR\otimes\CC.
  $$
  The isomorphism $\comp_{\ZZ/n,\BB}$ [(1.25)](#chunk-1.25) is induced by
  $$
    \exp(x/n)\colon \ZZ(1)_\BB \to \ZZ/n(1)\subset\CC^\times.
  $$
  The isomorphism $\comp_{\crisp,\DR}$ is induced by the inclusion $\QQ\subset\QQ_p$.
- [(M'5)](minus-three-points-mp5): Replace $\CC$ by $C$ in [(M5)](minus-three-points-m5).
- [(M6)](minus-three-points-m6) and [(M8)](minus-three-points-m8): $\ZZ(1)$ is of pure weight $-2$, and $\ZZ(1)_\DR$ is of pure Hodge filtration $-1$: the Hodge type is $(-1,-1)$.
:::

@@

::: {.trmenv #M7P title="(M7)P" latex="{(M7)P}"}
\oldpage{108}The data of $F_\infty$ for $E$ is equivalent to that of an involution of $P_\BB$ that is compatible with that of $M_\BB$.
Axiom [(AM2)](minus-three-points-am2) for $E$ is that this involution is induced by the $\CC$-antilinear involution of $P_{\BB,\CC}=\PP_{\DR,\CC}$ with fixed points $P_{\DR,\RR}$.
:::

::: {.trmenv #M8P title="(M8)P" latex="{(M8)P}"}
The data of the Hodge filtration of $E_\DR$ is equivalent to that of the intersection $F^0P_\DR$ of $F^0E_\DR$ over $P_\DR$, which is a torsor for $F^0M_\DR$: for $p>0$, $F^pE_\DR=F^pM_\DR$; for $p=0$, $F^0E_\DR$ is generated by $F^0P_\DR$; and for $p<0$, $F^pE_\DR$ is the sum of $F^0E_\DR$ and $F^pM_\DR$.
In other words, the data [(M8)](minus-three-points-m8) for $E_\DR$ is that of a reduction of the structure group of $P_\DR$, from $M_\DR$ to $F^0M_\DR$.

The data of [(M1)](minus-three-points-m1), [(M2)](minus-three-points-m2), $\comp_{\DR/\BB}$, and [(M8)](minus-three-points-m8) are equivalent to the data of a torsor $F^0P_\DR$ for $F^0M_\DR$ and, in the $M_\DR\otimes\CC=M_\BB\otimes\CC$ torsor $P_{\DR,\CC}$ which is induced by $F^0M_\DR\to M_\DR\otimes\CC$, a $\QQ$-structure $P_\BB\subset P_{\DR,\CC}$ that is an $M_\BB$ torsor.
:::

@@

We leave to the reader the task of translating axioms [(AM1)](minus-three-points-am1) to [(AM5)](minus-three-points-am5) for $P$.
We note only that [(AM1)](minus-three-points-am1) and [(AM4)](minus-three-points-am4) follow from the same axioms for $M$, and that [(AM2)](minus-three-points-am2) and [(AM5)](minus-three-points-am5) can be expressed at best as functoriality in $\CC$, cf. \ref{minus-three-points-chunk-1.14}.
::::
